A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-03-28 paper/paper.tex
Changes in paper Tested everything on quipu.

The SAFT-VR dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp}[n] &= \int \left(a_1(\xx) + \beta a_2(\xx)\right)n(\xx)d\xx
\end{align}
where $a_1$ and $a_2$ are the first two terms in a high-temperature
perturbation expansion and $\beta=1/k_BT$.  The first term ,$a_1$, is 
the mean-field dispersion interaction and the second term, $a_2$, describes the
effect of fluctuations resulting from compression of the fluid due
to the dispersion interaction itself. $a_2$ is approximated
using the local compressibility approximation (LCA), which
assumes the energy fluctuation is directly related to the
compressibility of a hard-sphere reference fluid\cite{barker1976liquid}.

@@

The second term, $a_2$, which describes the contribution to the free
energy associated with fluctuations is given by
\begin{align}
  a_2 &= \frac12 \epsilondisp
              K_{HS}(\etadisp(\xx)) \etadisp(\xx)
              \frac{\partial a_1}{\partial \etadisp(\xx)}
\end{align}
where $K_{HS}$ is the Percus-Yevick hard-sphere compressibility, given
by
\begin{align}
  K_{HS} &=
    \frac{(1 - \etadisp)^4}{1 + 4(\etadisp + \etadisp^2)}.
\end{align}

@@

From the Helmholtz free energy functional, we may obtain any other
thermodynamic functions we should choose.  The grand free energy
$\Phi$, which we ordinarily use in our computations, is obtained by
simply adding a chemical potential term:
\begin{equation}
  \Phi[n] = F[n] + \mu \int n(\xx) d\xx.
\end{equation}

@@

To estimate how the transition from vapor to liquid between the rods depends on their 
radius, we will again use this model and propose that 
the critical separation, $d_c$, will occur when the free energies are equal. This 
corresponds to equating the circumference of the stadium-shape (see 
top of Figure~\ref{fig:density-rods}) and the circumference of the two rods (see bottom
of Figure~\ref{fig:density-rods}), which results in
\begin{equation}
d_c = (\pi-2)r.\label{criticalseparation}
\end{equation}
The predicted and actual computed values for $d_c$ are shown in 
Table~\ref{criticalseparationtable}, and they match fairly well for the 
0.5~-~1.0~nm rods. We hypothesize that larger radii rods
would also follow this trend, but smaller rods are likely to have a different
relationship with the surrounding water molecules. This may explain why the critical
separation result for a radius of 0.3~nm does not match the prediction well.

@@

\begin{table}
\begin{tabular} {|c|c|c|c|}
\hline
$r$ (nm) & $d_c$ prediction (nm) & $d_c$ result (nm) & \% error \\
\hline
0.3 & 0.34 & 0.23 & 32\% \\
\hline
0.5 & 0.57 & 0.54 & 5.3\% \\
\hline
0.7 & 0.80 & 0.81 & 1.3\% \\
\hline
0.9 & 1.03 & 1.06 & 2.9\% \\
\hline
1.0 & 1.14 & 1.19 & 4.4\% \\
\hline
\end{tabular}
\caption{Calculations of the critical separation 
(see Equation~\ref{criticalseparation})
for the transition
from liquid to vapor between two hydrophobic rods.\needsworknow{Delete
this table and move some of the information into the text along with
prediction model and explanation of Figure~\ref{fig:rods-energy-vs-distance}}.}
\label{criticalseparationtable}
\end{table}

@@

We simulate four hydrophobic rods in water using the same radii as the two rods and
separation distances $d = 0$~nm to $d = 3$~nm for nearest neighbor rods. We placed
the rods in a 'square' arrangement and observed a similar behavior to the two rods
where vapor is present in between the rods(shown in Fig. \ref{fig:density-four-rods}
as 'before') until a critical separation where it becomes filled with water
(shown in Fig. \ref{fig:density-four-rods} as 'after'). For large $d$,
the rods behave like separate single rods.  Free energy per length
is plotted against rod separation in Fig. \ref{fig:four-rods-energy-vs-distance}.
From the slope we calculate the force per length to be [number].

The rounded square formed by the rods has a circumference estimated by
\begin{equation}
C_{rs} = 2 \pi r + 8r + 4d.
\end{equation}
We again equate this circumference with the total circumference of the four rods, which
results in a critical separation
\begin{equation}
d_{c4} = \frac{3 \pi - 4}{2} r \label{criticalfourrods}
\end{equation}

@@

Figure~\ref{fig:density-sphere} shows the density profile for several
hard sphere radii, plotted together with the results of the same
SPC/E molecular dynamics simulation shown in
Figure~\ref{fig:sphere-energy-vs-diameter}\cite{huang2001shs}.  The
agreement with simulation is quite reasonable.  The largest
disagreement involves the density at contact, which according to the
contact value theorem cannot agree, since the free energies do not
agree.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/sphere-density}
\end{center}
\caption{ Density profiles illustrating the transition from vapor 
to liquid water between the rods. The radius is 0.5~nm, the top figure is 
at a separation of 0.2~nm and the
bottom is 0.6~nm. Figures~\ref{fig:rods-energy-vs-distance}~and~\ref{fig:energy-rods} show
the energy for these and other separations.}
\label{fig:sphere-density}
\end{figure}

>>>>

The SAFT-VR dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp}[n] &= \int \left(a_1(\xx) + \beta a_2(\xx)\right)n(\xx)d\xx
\end{align}
where $a_1$ and $a_2$ are the first two terms in a high-temperature
perturbation expansion and $\beta=1/k_BT$.  The first term, $a_1$, is 
the mean-field dispersion interaction and the second term, $a_2$, describes the
effect of fluctuations resulting from compression of the fluid due
to the dispersion interaction itself. $a_2$ is approximated
using the local compressibility approximation (LCA), which
assumes the energy fluctuation is directly related to the
compressibility of a hard-sphere reference fluid\cite{barker1976liquid}.

@@

@@

From the Helmholtz free energy functional, we may obtain any other
thermodynamic functions we should choose.  The grand free energy
$\Omega$, which we ordinarily use in our computations, is obtained by
simply adding a chemical potential term:
\begin{equation}
  \Omega[n] = F[n] + \mu \int n(\xx) d\xx.
\end{equation}

@@

To estimate how the transition from vapor to liquid between the rods depends on their 
radius, we will again use this model and propose that 
the critical separation, $d_c$, will occur when the free energies are equal. This 
corresponds to equating the circumference of the stadium-shape (see 
top of Figure~\ref{fig:density-rods}) and the circumference of the two rods (see bottom
of Figure~\ref{fig:density-rods}), which results in
\begin{equation}
d_c = (\pi-2)r.\label{criticalseparation}
\end{equation}
The predicted and actual computed values for $d_c$ are shown graphically in 
Figure~\ref{fig:rods-energy-vs-distance}, and they match fairly well for the 
0.5~-~1.0~nm rods with differences between 1.3~-~5.5\%. We hypothesize that larger radii rods
would also follow this trend, but smaller rods are likely to have a different
relationship with the surrounding water molecules. This may explain why the critical
separation result for a radius of 0.3~nm does not match the prediction
well(32\% difference).

@@

@@

We simulate four hydrophobic rods in water using the same radii as our
two rods ($r = 0.3$~nm~-~$1.0$~nm)and
separation distances between $d = 0$~nm to $d = 3$~nm for nearest neighbor rods. We placed
the rods in a 'square' arrangement and observed a similar behavior as the two rods
where vapor is present in the space between the rods until a critical separation
where it becomes filled with water (shown in Fig. \ref{fig:density-four-rods}). 
For large $d$,
the rods behave like separate single rods.  Free energy per length
is plotted against rod separation in Figure \ref{fig:four-rods-energy-vs-distance}.
From the slope we calculate the force per length to be [number].

The rounded square formed by the area of the four rods and the space
inbetween them has a circumference estimated by
\begin{equation}
C_{rs} = 2 \pi r + 8r + 4d.
\end{equation}
We again equate this circumference with the total circumference of
only the four rods, which results in a critical separation of
\begin{equation}
d_{c4} = \frac{3 \pi - 4}{2} r. \label{criticalfourrods}
\end{equation}
We see in Figure \ref{fig:four-rods-energy-vs-distance} that the $r = 3.0$~nm
rod has the largest difference between simulation and prediction with
a 23\% difference, and better agreement for larger radii (between 0.5\% and 5.6\%).

@@

Figure~\ref{fig:density-sphere} shows the density profile for several
hard sphere radii, plotted together with the results of the same
SPC/E molecular dynamics simulation shown in
Figure~\ref{fig:sphere-energy-vs-diameter}\cite{huang2001shs}.  The
agreement with simulation is quite reasonable.  The largest
disagreement involves the density at contact, which according to the
contact value theorem cannot agree, since the free energies do not
agree.c
